The effect of climate forcing on numerical simulations of the Cordilleran Ice Sheet at the last glacial maximum

### 2013-06-13 cordillera-climate.tex
General clean-up in the todonotes

% Add to-do notes and enlarge right margin
\usepackage{todonotes}
\usepackage[left=54pt, top=81pt, textwidth=398pt, marginparsep=11pt, marginparwidth=93pt, textheight=625pt, footskip=30pt]{geometry}
\newcommand{\needref}[0]{\todo[color=cyan]{Reference needed}}
\newcommand{\arjen}[2][]{\todo[color=lime,{#1}]{\textbf{Arjen:} #2}}
\newcommand{\constantine}[2][]{\todo[color=yellow,{#1}]{\textbf{Constantine:} #2}}
\newcommand{\qiong}[2][]{\todo[color=pink,{#1}]{\textbf{Qiong:} #2}}
\newcommand{\ask}[2][]{\todo[color=lightgray,{#1}]{\textbf{Julien:} #2}}

@@

% ----------------------------------------------------------------------
\begin{abstract}
\todo{This abstract is preliminary. I will rewrite it when I am done with the rest}
We present an ensemble of numerical simulations of the former Cordilleran ice sheet forced by present-day climatologies from five different datasets. Surface air temperature and precipitation rates from the NCEP/NCAR reanalysis, the Climate Forecast System Reanalysis, the ERA-Interim reanalysis and the North American Regional Reanalysis are used to force the ice sheet model PISM. Glacial conditions are mimicked by applying constant temperature offsets over the entire modelling domain, and a positive degree-day model is used to compute surface mass balance. The modelled ice sheet appears highly sensitive to the choice of climate forcing. Best results were obtained from the North American Regional Reanalysis, a high-resolution regional reanalysis which assimilates precipitation.
\end{abstract}
% ----------------------------------------------------------------------

>>>>

% Add to-do notes and enlarge right margin
\usepackage{todonotes}
\usepackage[left=54pt, top=81pt, textwidth=398pt, marginparsep=11pt, marginparwidth=93pt, textheight=625pt, footskip=30pt]{geometry}
\newcommand{\needref}[0]{\todo[color=cyan]{Reference needed}}
\newcommand{\arjen}[2][]{\todo[color=lime,{#1}]{\textbf{Arjen:} #2}}
\newcommand{\constantine}[2][]{\todo[color=yellow,{#1}]{\textbf{Constantine:} #2}}
\newcommand{\qiong}[2][]{\todo[color=pink,{#1}]{\textbf{Qiong:} #2}}
\newcommand{\julien}[2][]{\todo[color=lightgray,{#1}]{\textbf{Julien:} #2}}

@@

% ----------------------------------------------------------------------
\begin{abstract}
\julien[noline]{The abstract is preliminary.}
We present an ensemble of numerical simulations of the former Cordilleran ice sheet forced by present-day climatologies from five different datasets. Surface air temperature and precipitation rates from the NCEP/NCAR reanalysis, the Climate Forecast System Reanalysis, the ERA-Interim reanalysis and the North American Regional Reanalysis are used to force the ice sheet model PISM. Glacial conditions are mimicked by applying constant temperature offsets over the entire modelling domain, and a positive degree-day model is used to compute surface mass balance. The modelled ice sheet appears highly sensitive to the choice of climate forcing. Best results were obtained from the North American Regional Reanalysis, a high-resolution regional reanalysis which assimilates precipitation.
\end{abstract}
% ----------------------------------------------------------------------

### 2013-06-13 sections/climate.tex
General clean-up in the todonotes

\arjen{It is an unusual style to start each sentence with Figure X shows... We should rephrase this later-on.} Figure~\ref{fig:prec} shows the spatial distribution of DJF and JJA precipitation rates in WorldClim. Coastal regions generally receive much more precipitation than inland ones as a result of the continuous topographic barrier formed by the Boundary Ranges, the Coast Mountains and the Cascades\arjen{orographic precip!}. From a glacier mass-balance point of view, this contrast is made even stronger by the difference in timing of the precipitation peak through the year. Although coastal regions experience most precipitation as snow during the accumulation season, inland regions experience dry winters and most of the precipitation falls as rain during the summer months.

@@

The density of weather stations used by WorldClim in the Northern American Cordillera is highly inhomogeneous. If good coverage exists in the southern parts of our modelling domain, several hundred kilometres can separate neighbouring stations in the north \citep{data:worldclim}. A second drawback of the dataset in the scope of this study is the lack of data across oceans, particularly on the Pacific continental shelf where glaciers have advanced during the last glacial cycle\needref.

@@

Air surface temperature and precipitation rates \arjen{do you really mean rate as in mm/time unit?} from the WorldClim data, the NCEP/NCAR, ERA-Interim, CFSR and NARR reanalyses were re-projected to Canadian Atlas Lambert conformal conic projection (EPSG code~3978) \arjen{is this general knowledge? I am lost...} and bilinearly interpolated to the model resolution of 10\,km. In addition, WorldClim data was extrapolated to oceans using a nearest-neighbour approach. These last two steps are not presented on figures~\ref{fig:temp} and~\ref{fig:prec}.

For the CFSR, an alternative forcing was prepared by smoothing the precipitation field in order to correct for artefacts seen in figure~\ref{fig:prec}. This was done by averaging data locally in a circular neighbourhood of 7\,pixels in diameter prior to re-projection. Yet we will show that this smoothing has very limited effect on the ice sheet model outcome. This preprocessing steps were made in GRASS~GIS using scripts available on the first author's website\arjen{I am unsure if this is the best way t do it? Were you using principles described by others then you should refer to the litterature instead.}.

>>>>

Figure~\ref{fig:prec} shows the spatial distribution of DJF and JJA precipitation rates in WorldClim. Coastal regions generally receive much more precipitation than inland ones as a result of the continuous orographic barrier formed by the Boundary Ranges, the Coast Mountains and the Cascades. From a glacier mass-balance point of view, this contrast is made even stronger by the difference in timing of the precipitation peak through the year. Although coastal regions experience most precipitation as snow during the accumulation season, inland regions experience dry winters and most of the precipitation falls as rain during the summer months.

@@

The density of weather stations used by WorldClim in the Northern American Cordillera is highly inhomogeneous. If good coverage exists in the southern parts of our modelling domain, several hundred kilometres can separate neighbouring stations in the north \citep{data:worldclim}. A second drawback of the dataset in the scope of this study is the lack of data across oceans, particularly on the Pacific continental shelf where glaciers have advanced during the last glacial cycle \citep{jackson-clague-1991}.

@@

Air surface temperature and precipitation from the WorldClim data, the NCEP/NCAR, ERA-Interim, CFSR and NARR reanalyses were re-projected to Canadian Atlas Lambert conformal conic projection (EPSG code~3978) and bilinearly interpolated to the model resolution of 10\,km. In addition, WorldClim data was extrapolated to oceans using a nearest-neighbour approach. These last two steps are not presented on figures~\ref{fig:temp} and~\ref{fig:prec}.

For the CFSR, an alternative forcing was prepared by smoothing the precipitation field in order to correct for artefacts seen in figure~\ref{fig:prec}. This was done by averaging data locally in a circular neighbourhood of 7\,pixels in diameter prior to re-projection. Yet we will show that this smoothing has very limited effect on the ice sheet model outcome. These preprocessing steps were made in GRASS~GIS.

### 2013-06-13 sections/intro.tex
General clean-up in the todonotes

Another method consist in using atmospheric fields produced in uncoupled palæo-climate simulations \citep{huybrechts-tsiobbel-1996}. To produce a time-dependant forcing, the GCM output is interpolated in time between different simulations \citep{charbit-etal-2002} or weighted based on ice-core records \citep{marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}. These palæo-climate simulations use ice sheet reconstructions such as the ICE-4G \citep{peltier-1994} as topographic boundary condition, which results in subsequently modelled ice cover consistent with these reconstructions.
\ask[noline]{I am thinking whether I should extend this literature list to papers not concerning North America?}

@@

Here we use a Parallel Ice Sheet Model \citep[PISM,][]{web:pism} to simulate the entire Cordilleran Ice Sheet at the Last Glacial Maximum (LGM). We force our model with multiple climate datasets and compare our results to the mapped LGM ice sheet margin from \citet{kleman-etal-2010}. See \citet{quiquet-etal-2012} for a similar comparison on Greenland. To avoid any dependence on ice-sheet reconstructions and allow high spatial resolution in our climate forcing, we use present-day climate data and given temperature offsets in a manner similar to \citet{bintanja-etal-2005}. To our knowledge, this is the first modelling study which specifically focus on the Cordilleran Ice Sheet since the one by \citep{robert-1991}, who focused specifically on its southern margin.

>>>>

Another method consist in using atmospheric fields produced in uncoupled palæo-climate simulations \citep{huybrechts-tsiobbel-1996}. To produce a time-dependant forcing, the GCM output is interpolated in time between different simulations \citep{charbit-etal-2002} or weighted based on ice-core records \citep{marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}. These palæo-climate simulations use ice sheet reconstructions such as the ICE-4G \citep{peltier-1994} as topographic boundary condition, which results in subsequently modelled ice cover consistent with these reconstructions.
\julien[noline]{Should this literature list be extended to studies not covering North America?}

@@

Here we use a Parallel Ice Sheet Model \citep[PISM,][]{web:pism} to simulate the entire Cordilleran Ice Sheet at the Last Glacial Maximum (LGM). We force our model with multiple climate datasets and compare our results to the mapped LGM ice sheet margin from \julien{I will try to find an outline that is not cropped over Alaska instead.}\citet{kleman-etal-2010}. See \citet{quiquet-etal-2012} for a similar comparison on Greenland. To avoid any dependence on ice-sheet reconstructions and allow high spatial resolution in our climate forcing, we use present-day climate data and given temperature offsets in a manner similar to \citet{bintanja-etal-2005}. To our knowledge, this is the first modelling study which specifically focus on the Cordilleran Ice Sheet since the one by \citep{robert-1991}, who focused specifically on its southern margin.

### 2013-06-13 sections/model.tex
General clean-up in the todonotes

Monthly temperature and precipitation are converted to surface accumulation and ablation by a temperature-index (positive degree-day) model. Accumulation is computed as the amount of precipitation that falls when temperature is below freezing\ask{Is this true? What is the current threshold in stable 0.5?}. Ablation is computed from the number of positive degree-days, defined as the integral of temperatures above 0\,\degree\,C in one year. Positive degree-days are calculated by the \citet{calov-greve-2005} formula to accounts for day-to-day temperature variability by assuming a normal distribution of variance $\sigma=5\degree C$ along a central (input) value $T$:

@@

Ice flow velocities are computed by a heuristic \arjen{not sure what this expresses!!} where the Shallow Shelf Approximation (SSA) is used as a ``sliding law'' to a polythermal Shallow Ice Approximation (SIA) \citep{bueler-brown-2009}.

\ask[inline]{\emph{Should I include the SSA+SIA \citet{bueler-brown-2009} formula?}

	Depth-averaged velocities $\vec{v}$ are computed as a weighted sum of SIA velocities $\vec{v_{\mathrm{SIA}}}$ and SSA velocities $\vec{v_{\mathrm{SSA}}}$ by

@@

	\begin{equation}
		f(|\vec{v_{\mathrm{SSA}}}|) = 1
		- \frac{2}{\pi}\arctan{\frac{|\vec{v_{\mathrm{SSA}}}|^2}{100^2}}.
	\end{equation}
}

\arjen[inline]{unless variations in these determine your best fit input dataset, I would not include them in this paper.}

@@

With $\phi$ in degrees and $z$ in meters above present sea level. The glacier bed topography responds to the ice load following a regional isostasy model with flexure and relaxation \citep{lingle-clark-1985}\todo{Check \citet{lingle-clark-1985}}.

@@

In this study we model glacial inception and growth of the Cordilleran ice sheet to a state approaching its last glacial maximum. We mimic glacial conditions by applying constant temperature offsets homogeneously over the entire domain. Each simulation starts from ice-free conditions and is run for 10,000 years, a time interval representative of the rapid build-up of the last Cordilleran Ice Sheet from nearly ice-free condition to last glacial maximum extent \needref. Our simulations were run in parallel on 32 cores at the Swedish National Supercomputing Center.

\arjen[inline]{Check Stroeven et al. (2010): "Radiocarbon ages from wood and fossils underlying McConnell till indicate that the McConnell glaciation began after 29,600  300 14C yr BP (Matthews et al., 1990), but glaciers did not advance into major valleys before 26,350  280 14C yr BP (Jackson and Harington, 1991), and a fully developed CIS did not exist until after 23,900  1140 14C yr BP (Lowdon and Blake,1981). Published radiocarbon ages indicate that the St. Elias lobe of the CIS was in retreat by 13,660  180 14C yr BP (Rampton, 1971) and the Selwyn lobe by 12,590  120 14C yr BP (Ward, 1989) (Fig. 2)." and references therein.}

>>>>

Monthly temperature and precipitation are converted to surface accumulation and ablation by a temperature-index (positive degree-day) model. Ice accumulation is equal to precipitation when temperature is below 0\,\degC, and decreases linearly with temperature between 0 and 2\,\degC. Ice ablation is computed from the number of positive degree-days, defined as the integral of temperatures above 0\,\degC in one year. Positive degree-days are calculated by the \citet{calov-greve-2005} formula to accounts for day-to-day temperature variability by assuming a normal distribution of variance $\sigma=5\degC$ along a central (input) value $T$:

@@

Ice flow velocities are computed by a heuristic the Shallow Shelf Approximation (SSA) is used as a ``sliding law'' to a polythermal Shallow Ice Approximation (SIA) \citep{bueler-brown-2009}. Depth-averaged velocities $\vec{v}$ are computed as a weighted sum of SIA velocities $\vec{v_{\mathrm{SIA}}}$ and SSA velocities $\vec{v_{\mathrm{SSA}}}$ by

@@

\begin{equation}
	f(|\vec{v_{\mathrm{SSA}}}|) = 1
	- \frac{2}{\pi}\arctan{\frac{|\vec{v_{\mathrm{SSA}}}|^2}{100^2}}.
\end{equation}

@@

With $\phi$ in degrees and $z$ in meters above present sea level. The glacier bed topography responds to the ice load following a regional isostasy model with flexure and relaxation \citep{lingle-clark-1985}.

@@

In this study we model glacial inception and growth of the Cordilleran ice sheet to a state approaching its last glacial maximum. We mimic glacial conditions by applying constant temperature offsets homogeneously over the entire domain. Each simulation starts from ice-free conditions and is run for 10,000 years, a time interval representative of the rapid build-up of the last Cordilleran Ice Sheet from nearly ice-free condition to last glacial maximum extent \citep{clague-1989}. Our simulations were run in parallel on 32 cores at the Swedish National Supercomputing Center.

### 2013-06-13 sections/results.tex
General clean-up in the todonotes

\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=13cm]{cordillera-climate-extent}
	\end{center}
	\qiong[inline]{Move ticks on the colorbar. The 1 is confusing.}
	\caption{Extent of ice cover after 10\,kyr as a function of applied temperature offsets for each climate forcing.}
	\label{fig:extent}
\end{figure*}

@@

\begin{figure}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=9cm]{cordillera-climate-ivolarea}
	\end{center}
	\todo[inline]{Use black-and-white-proof markers and highlight the ``best'' runs.}
	\caption{Total ice volume and glaciated area after 10\,kyr as a function of temperature offset and climate forcing.}
	\label{fig:ivolarea}
\end{figure}

@@

\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=13cm]{cordillera-climate-best}
	\end{center}
	\todo[inline]{The must would be to overlay geomorphic ice extent.}
	\caption{Ice surface topography (black contours every 1000\,m) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr using temperature offsets that lead to similar areas of ice cover for each climate forcing.}
	\label{fig:best}
\end{figure*}

To compare modelled ice sheets of similar size, we selected for each climate forcing the simulation that lead to a glaciated area closest to an arbitrary target of $4\,\times10^6\,\unit{km^2}$. These qualitative ``best'' runs are presented in Figure~\ref{fig:best} along the associated temperature offset values. We observe that although similar in size, the selected ice sheets are different in shape and flow patterns. More specifically, NCEP/NCAR, CFSR and ERA-Interim ice sheets cover large territories in Yukon and Alaska that where shown to have remained ice-free for at least several glacial cycles.\needref

>>>>

\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=13cm]{cordillera-climate-extent}
	\end{center}
	\caption{Extent of ice cover after 10\,kyr as a function of applied temperature offsets for each climate forcing.}
	\label{fig:extent}
\end{figure*}

@@

\begin{figure}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=9cm]{cordillera-climate-ivolarea}
	\end{center}
	\julien[inline]{I will make the figure readable in black-and-white and highlight markers corresponding to the ``best'' runs.}
	\caption{Total ice volume and glaciated area after 10\,kyr as a function of temperature offset and climate forcing.}
	\label{fig:ivolarea}
\end{figure}

@@

\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=13cm]{cordillera-climate-best}
	\end{center}
	\julien[inline]{I will to overlay the geomorphologic LGM ice margin so that it becomes easier to see which runs match it better.}
	\caption{Ice surface topography (black contours every 1000\,m) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr using temperature offsets that lead to similar areas of ice cover for each climate forcing.}
	\label{fig:best}
\end{figure*}

To compare modelled ice sheets of similar size, we selected for each climate forcing the simulation that lead to a glaciated area closest to an arbitrary target of $4\,\times10^6\,\unit{km^2}$. These qualitative ``best'' runs are presented in Figure~\ref{fig:best} along the associated temperature offset values. We observe that although similar in size, the selected ice sheets are different in shape and flow patterns. More specifically, NCEP/NCAR, CFSR and ERA-Interim ice sheets cover large territories in Yukon and Alaska that where shown to have remained ice-free for at least several glacial cycles\citep{dukrodkin-1999,kaufman-manley-2004}.
